Clustering via Generalized Energy Statistics

### 2018-02-09 energy_clustering/icml18/energy_clustering.tex
energy icml18

\documentclass[twoside]{article}
\usepackage{aistats2018}
 % If your paper is accepted, change the options for the package
% aistats2018 as follows:
%
%\usepackage[accepted]{aistats2018}
%
% This option will print headings for the title of your paper and
% headings for the authors names, plus a copyright note at the end of
% the first column of the first page.

\usepackage[compress,round]{natbib}
\usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm,bbm}
\usepackage{mathtools}
\mathtoolsset{showonlyrefs}
\usepackage[inline]{enumitem}
\usepackage[inline]{enumitem}
\usepackage[pdftex]{graphicx}
\graphicspath{{./figs/}}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{multirow}
\usepackage{colortbl}
\usepackage{array,booktabs}
\usepackage{xcolor}
\usepackage{makecell}
\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}

@@

@@

% If your paper is accepted and the title of your paper is very long,
% the style will print as headings an error message. Use the following
% command to supply a shorter title of your paper so that it can be
% used as headings.
%
%\runningtitle{I use this title instead because the last one was very long}

% If your paper is accepted and the number of authors is large, the
% style will print as headings an error message. Use the following
% command to supply a shorter version of the authors names so that
% they can be used as headings (for example, use only the surnames)
%
%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}

\twocolumn[

\aistatstitle{Clustering via Generalized Energy Statistics}

\aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }

\aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3 }

%\aistatsauthor{Guilherme Fran\c{c}a \And  Joshua T. Vogelstein}

%\aistatsaddress{ Johns Hopkins University }

]

\begin{abstract}
Energy statistics introduces the notion of potential
energy between probability distributions, in  
close analogy to Newton's gravitational
potential in physics. We propose a principled approach to the
\mbox{clustering} problem based on energy statistics theory.
Our mathematical formulation establishes connection to kernel methods,
leading to a quadratically constrained quadratic program.
To obtain local solutions of such an optimization problem, which is NP-hard, 
we introduce an iterative
algorithm based
on Hartigan's method, which has the same computational cost but offers
several advantages compared to the 
well-known kernel $k$-means, based on Lloyd's heuristic.
We provide carefully designed numerical experiments showing
that our method is more flexible and outperforms
kernel $k$-means, spectral clustering,
standard $k$-means and Gaussian mixture models in a variety of settings,
specially in high \mbox{dimensions}. 
\end{abstract}

@@

Energy statistics \citep{Szkely2013}
provides a hypothesis test for equality of 
distributions, which is achieved 
under the minimum of the statistical potential energy. 
When probability distributions are different the 
statistical potential energy diverges as sample size increases, while tends 
to a nondegenerate limit distribution when probability
distributions are equal. 
The test statistic has compact representation
in terms of expectations of pairwise distances, providing
straightforward empirical estimates. Energy statistics
has been applied to several goodness-of-fit 
hypothesis tests, multi-sample tests of equality of distributions, 
analysis of variance \citep{RizzoVariance}, nonlinear dependence tests through
distance covariance and distance correlation, which generalizes the Pearson
correlation coefficient, and hierarchical clustering \citep{RizzoClustering} 
by extending Ward's method of minimum variance. Moreover, an application of 
energy statistics to clustering in Euclidean spaces was 
considered \citep{Kgroups}.  
We refer to \citep{Szkely2013} for an overview.

Distance covariance was further generalized from Euclidean 
to metric spaces of strong negative type \citep{Lyons}. Furthermore, 
the missing link between energy distance based tests and kernel 
based tests has 
been recently resolved \citep{Sejdinovic2013}, where a unifying framework
establishing an equivalence between generalized energy distances to maximum
mean discrepancies (MMD), which are distances between embeddings of 
distributions in reproducing kernel Hilbert spaces (RKHS), was established. 
This equivalence immediately relates energy statistics to
kernel methods often used in machine learning, and form the basis 
of our approach.

@@

To account for nonlinearities, kernel methods were introduced 
\citep{Smola,Girolami}. A mercer kernel \citep{Mercer} is used to implicitly
map data points to a RKHS, then clustering can be performed in the associated
Hilbert space by exploiting its inner product. However, the kernel choice remains 
the biggest challenge since there is no principled theory to construct a kernel
for a given dataset, and usually kernels introduce hyperparameters that 
need to be carefully chosen.
The well-known kernel $k$-means optimization problem is nothing but $k$-means 
in the feature space \citep{Girolami}. Furthermore, kernel $k$-means algorithm
\citep{Dhillon2,Dhillon} is still based on Loyd's heuristic \citep{Lloyd}
of grouping points that are closer to a cluster center, now
in feature space. 
We refer the reader to \citep{Filippone} for a survey of clustering
methods.

Although clustering from energy statistics in Euclidean spaces was considered
in \citep{Kgroups}, the precise optimization problem behind this approach
remains elusive, as well as the connection with kernel methods.
The main theoretical contribution of this paper is to fill this gap.
Since the statistical potential energy is minimum when
distributions are equal, the principle behind clustering is to maximize 
the statistical energy,  enforcing probability distributions associated to 
each cluster to be different from one another. We provide a precise 
mathematical formulation to this statement, leading to a quadratically 
constrained quadratic program (QCQP) in the associated RKHS. Our results
immediately establish the connection between energy statistics based
clustering and kernel methods. Moreover, we show that that this 
QCQP is equivalent to kernel $k$-means optimization problem.

Our main 
algorithmic contribution is to use Hartigan's method \citep{Hartigan} to 
find local solutions of the above mentioned QCQP, which is NP-hard in general.
Hartigan's method was also used in \citep{Kgroups}, but without any
connection to kernels. More importantly, the 
advantages of Hartigan's
over Lloyd's method was already demonstrated 
in some simple settings
\citep{Telgarsky,Slonin}, but apparently this method did not receive 
the deserved attention. To the 
best of our knowledge, Hartigan's method was not previously 
employed together with kernel methods. 
We provide a fully kernel based Hartigan's algorithm for clustering,
where the kernel is fixed by energy statistics. 
We make clear the advantages of this proposal versus 
Lloyd's method, which kernel $k$-means is based upon and will also be used 
to solve our QCQP. We show that both algorithms  have the same
time complexity, however, Hartigan's method in kernel spaces is superior. 
Furthermore, in the examples considered in this paper, it 
also provides equal or better performance than spectral clustering,
which in fact solves a relaxed version of our QCQP.
Our numerical results provide compelling evidence that 
Hartigan's method applied to energy statistics based clustering
is more accurate and robust than 
kernel $k$-means. Moreover, we illustrate the
flexibility of energy clustering,  
showing that it is able to perform accurately on data coming from 
very different distributions, contrary to $k$-means and GMM for instance.
More specifically, the proposed method performs 
closely to $k$-means and GMM on normally distributed data, however,
it is significantly better on other settings. 
Its superiority in high dimensions is striking, being more accurate 
than $k$-means and GMM even on Gaussian settings.

@@

In this section we introduce the main concepts from (generalized) energy
statistics \citep{Szkely2013,Lyons} and its relation to 
RKHS \citep{Sejdinovic2013}
which form the basis of our work.

Consider random vectors $X_i$ living in an arbitrary space
$\mathcal{X}$ of \emph{negative type}, which means that $\mathcal{X}$
is endowed with a \emph{semimetric} 
$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$ satisfying
$\sum_{i,j=1}^n c_i c_j \rho(X_i, X_j) \le 0$,
where $c_i \in \mathbb{R}$ and
$\sum_{i=1}^n c_i = 0$. 
Let $X,X' \stackrel{iid}{\sim} P$ and 
$Y,Y' \stackrel{iid}{\sim} Q$, where $P$ and $Q$ are cumulative
distribution functions with finite first moments, and 
$X,X',Y,Y' \in \mathcal{X}$. 
The \emph{generalized energy distance} between $P$ and $Q$ is
given by 
\begin{equation}
\label{eq:energy3}
\Energy(P, Q) \equiv 2 \E \rho(X,Y) - \E \rho(X, X') - \E \rho(Y,Y').
\end{equation} 
This quantity is nonnegative, $\Energy(P,Q) \ge 0$, and 
$\Energy^{1/2}$ is
a metric on the space of distributions. Energy distance provides
a characterization of equality of distributions.
For instance, the \emph{standard
energy distance} \citep{Szkely2013} in Euclidean spaces uses
the semimetric
\begin{equation}
\label{eq:rho_standard}
\rho_\alpha(X,Y) = \| X - Y\|^\alpha
\end{equation} 
where $0< \alpha \le 2$ and $\| \cdot \|$ is the
Euclidean norm in $\mathcal{X}=\mathbb{R}^D$. In this case $\Energy(P,Q)$
is rotationally invariant. For $0<\alpha<2$ we have
$\Energy(P,Q) = 0$ if and only if $P=Q$. However, for $\alpha=2$
we get $\Energy(P,Q) = 2\| \E(X) - \E(Y) \|^2$, thus
$\Energy(P,Q)=0$ does 
not imply equality of distributions but only equality
of the means.

Consider a sample $\mathbb{X} = \{ x_1,\dotsc, x_n \}$ 
from $k$ unknown distributions $\{ P_j \}_{j=1}^k$,
with points lying on a space of negative type,
$x_i \in \mathcal{X}$.
Let $\mathbb{X} = \bigcup_{j=1}^k \C_j$ be a disjoint
partition, i.e. $\C_i \cap \C_j = \emptyset$.
Each expectation in the generalized energy distance
can be empirically estimated with the aid of the
function
\begin{equation}
\label{eq:g_def}
g (\C_i, \C_j) \equiv 
\dfrac{1}{n_i n_j}
\sum_{x \in \C_i} 
\sum_{y \in \C_j} \rho(x, y) ,
\end{equation}
where $n_i = |\C_i|$ denotes the number of elements in partition
$\C_i$. 
Define the \emph{within energy dispersion} as
\begin{equation}
\label{eq:within}
W \equiv
\sum_{j=1}^{k} \dfrac{n_j}{2} g(\C_j, \C_j),
\end{equation}
and the \emph{between-sample energy statistic} as
\begin{equation}
\label{eq:between}
S \equiv \hspace{-1em}
\sum_{1 \le  i < j \le k } \hspace{-.5em} \dfrac{n_i n_{j}}{2 n} \left[
2 g(\C_i, \C_j) - 
g(\C_i, \C_i) - 
g(\C_j, \C_j)
\right],
\end{equation}
where $n = \sum_{j=1}^k n_j$.
A given point $x_i$ belongs to partition $\C_j$
if and only if $x_i \sim P_j$. 
The quantity $S$ is
a test statistic for equality of distributions
\citep{Szkely2013}.
When the sample size is large enough, $n\to \infty$,
under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$ we have that
$S\to 0$, 
and under
the alternative hypothesis $H_1: P_\ell \ne P_j$ for at least two $\ell\ne j$, 
we have that $S \to \infty$.

Let $\HH_\kk$ be a Hilbert space of real-valued functions
over $\mathcal{X}$ with an associated kernel
$\kk : \mathcal{X} \times \mathcal{X} \to 
\mathbb{R}$, which is a symmetric and positive definite function, i.e. 
$\kk(x_i,x_j) = \kk(x_j,x_i)$ and 
$\sum_{i,j=1}^n c_i c_j \kk(x_i, x_j) \ge 0$, or equivalently,
if $G$ is the Gram matrix with entries $G_{ij} = \kk(x_i,x_j)$
then $G = G^\top$ and $v^\top G v \ge 0$ for any $v \in \mathbb{R}^n$.
For every $x \in \mathcal{X}$ there exists 
$h_x \equiv \kk(\cdot,x) \in \Hk$ such that  $\langle h_x, f \rangle = f(x)$
for any function $f \in \Hk$. Thus, 
$\langle h_x, h_y \rangle = \kk(x,y)$.
Conversely \citep{Aronszajn},  
for every symmetric
positive definite function $\kk: \mathcal{X}\times \mathcal{X} \to
\mathbb{R}$ there is a Hilbert space $\Hk$ with reproducing
kernel $\kk$, with a 
\emph{feature map} $\varphi: x \mapsto \kkk_x \in \Hk$ such
that $\langle \varphi(x), \varphi(y) \rangle = \kk(x, y)$.

Define the embedding of a probability measure
$P \mapsto \kkk_P \in \Hk$ through
$\kkk_P \equiv \int \kk( \, \cdot \,, x)  d P(x)$. 
The distance between two probability measures, 
called maximum mean discrepancy (MMD), is thus given by
\begin{equation}
\label{eq:mmd}
\gamma_\kk(P,Q) \equiv \| \kkk_P - \kkk_Q \|_{\Hk},
\end{equation}
which can also be written as \citep{Gretton2012}
\begin{equation}\label{eq:mmd2}
\gamma_\kk^2(P,Q) = \E \kk(X,X') + \E \kk(Y,Y') - 2 \E \kk(X, Y)
\end{equation}
where $X,X' \stackrel{iid}{\sim} P$ and $Y,Y'\stackrel{iid}{\sim} Q$.
The equality between \eqref{eq:mmd} and \eqref{eq:mmd2}
gives $\langle \kkk_P, \kkk_Q \rangle_{\Hk} = \E \, \kk(X, Y)$,
thus in practice we can estimate the inner product between  
embedded distributions by averaging the kernel function over sampled data.

The following important result shows that semimetrics of negative
type and symmetric positive definite kernels are closely related
\citep{Berg1984}. Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$,
and $x_0 \in \mathcal{X}$ be an arbitrary but fixed point.
Define
\begin{equation}
\label{eq:kernel_semimetric}
\kk(x,y) \equiv 
\tfrac{1}{2} \left[  \rho(x,x_0) + \rho(y,x_0) - \rho(x,y)\right].
\end{equation}
Then, it can be shown that 
$\kk$ is positive definite if and only if $\rho$ is a semimetric
of negative type.
We have a family of kernels, one for each choice of $x_0$. Conversely,
if $\rho$ is a semimetric of negative type and $\kk$ is a kernel in this
family, then 
\begin{equation}
\label{eq:gen_kernel}
\begin{split}
\rho(x,y) &= \kk(x,x) + \kk(y,y) -2\kk(x,y) \\
&=  \| \kkk_x - \kkk_y \|^2_{\Hk}
\end{split}
\end{equation}
and the canonical feature map 
$\varphi: x \mapsto \kkk_x$ is injective \citep{Sejdinovic2013}.
When these conditions are met we say that the kernel $\kk$ 
generates the semimetric $\rho$. 
If two different kernels generate the same $\rho$ they are
said to be equivalent kernels.
Now we can state the equivalence between energy distance and
inner products on RKHS, which is one of the main results of
\citep{Sejdinovic2013}. If $\rho$ is a semimetric
of negative type and $\kk$ a kernel that generates $\rho$, then
replacing \eqref{eq:gen_kernel} into
\eqref{eq:energy3}, and using \eqref{eq:mmd2}, yields
\begin{equation}
\label{eq:Erho}
\begin{split}
\tfrac{1}{2}\Energy(P, Q) &= 
\E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y) \\ 
&= \gamma_\kk^2(P,Q) .
\end{split}
\end{equation}
Therefore, we can compute the energy distance using the inner product
of $\Hk$.

@@

This section contains our main theoretical results where 
we formulate an optimization problem for clustering 
based on energy statistics in the RKHS.
The proofs are contained in supplementary material.

Due to the energy test statistic for equality of distributions,
the obvious
criterion for clustering data is to 
maximize $S$ which makes 
each cluster as different
as possible from the other ones.
In other words, given a set of points coming from different probability
distributions, the test statistic $S$ should attain a maximum when 
each point is correctly
classified as belonging to the cluster associated to its probability
distribution.
The following 
straightforward result
shows that maximizing $S$ is, however, equivalent to minimizing
$W$ which has a more convenient form.

\begin{lemma}
\label{th:minimize}
Let $\mathbb{X} = \{x_1,\dotsc,x_n\}$ where each data point
$x_i$ lives in a space $\mathcal{X}$ of negative type. 
For a fixed integer $k$,
the partition
$\mathbb{X} = \bigcup_{j=1}^k \C_j$, where $\C_i \cap C_j = \emptyset$ for
all $i\ne j$, maximizes the between-sample statistic $S$, defined
in \eqref{eq:between}, if and only if
\begin{equation}
\label{eq:minimize}
\min_{\C_1,\dotsc,C_k  } W(
\C_1, \dotsc, \C_k),
\end{equation}
where the within energy dispersion $W$ is defined by \eqref{eq:within}.
\end{lemma}

>>>>

%%%%%%%% ICML 2018 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\graphicspath{{./figs/}}
\usepackage{subfigure}
\usepackage{booktabs} % for professional tables

\usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm,bbm}
\usepackage{mathtools}

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2018} with \usepackage[nohyperref]{icml2018} above.
\usepackage{hyperref}
\usepackage[inline]{enumitem}
\usepackage{multirow}
\usepackage{colortbl}
\usepackage{array,booktabs}
\usepackage{xcolor}
\usepackage{makecell}
\usepackage{pifont}

% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
%\usepackage{icml2018}

% If accepted, instead use the following line for the camera-ready submission:
\usepackage[accepted]{icml2018}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Clustering via Generalized Energy Statistics}

@@

\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}

@@

\twocolumn[
\icmltitle{Clustering via Generalized Energy Statistics}

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the icml2018
% package.

% List of affiliations: The first argument should be a (short)
% identifier you will use later to specify author affiliations
% Academic affiliations should list Department, University, City, Region,
% Country
% Industry affiliations should list Company, City, Region, Country

% You can specify symbols, otherwise they are numbered in order.
% Ideally, you should not use this facility. Affiliations will be numbered
% in order of appearance and this is the preferred way.
\icmlsetsymbol{equal}{*}

\begin{icmlauthorlist}
\icmlauthor{Guilherme Fran\c ca}{to}
\icmlauthor{Joshua Vogelstein}{to}
\end{icmlauthorlist}

\icmlaffiliation{to}{Johns Hopkins University}

\icmlcorrespondingauthor{Guilherme Fran\c ca}{guifranca@jhu.edu}
\icmlcorrespondingauthor{Joshua Vogelstein}{jovo@jhu.edu}

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{Clustering, Energy Statistics}

\vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column
% listing the affiliations and the copyright notice.
% The command takes one argument, which is text to display at the start of the
% footnote.
% The \icmlEqualContribution command is standard text for equal contribution.
% Remove it (just {}) if you do not need this facility.

\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
%\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the
%standard text.

\begin{abstract}
Energy statistics introduces the notion of potential
energy between probability distributions, in  
close analogy to Newton's gravitational
potential in physics. In this paper, we propose a principled approach to 
clustering based on energy statistics theory.
Our mathematical formulation establishes connection to kernel methods,
leading to a quadratically constrained quadratic program in the associated
feature space.
To obtain local solutions of such an NP-hard optimization problem,
we introduce an iterative algorithm based
on Hartigan's method. This algorithm has the same computational 
cost as kernel $k$-means but offers
several advantages. 
We provide carefully designed numerical experiments illustrating
that the proposed method is more flexible and outperforms
kernel $k$-means, spectral clustering,
standard $k$-means and Gaussian mixture models in a variety of settings,
specially in high dimensions. We also employ the method 
in a real data application describing shapes of neural synapses.
\end{abstract}

@@

Energy statistics \citep{Szkely2013}
provides a hypothesis test for equality of 
distributions, which is achieved 
under the minimum of a ``statistical energy''. 
When probability distributions are different, this
statistical energy diverges as sample size increases. On the other hand,
it tends to a nondegenerate limit distribution when probability
distributions are equal.
The test statistic has a compact representation
in terms of expectations of pairwise distances, providing
straightforward empirical estimates. Energy statistics
has been applied to several goodness-of-fit 
hypothesis tests, multi-sample tests of equality of distributions, 
analysis of variance \citep{RizzoVariance}, nonlinear dependence tests through
distance covariance and distance correlation, which generalizes the Pearson
correlation coefficient, and hierarchical clustering \citep{RizzoClustering} 
by extending Ward's method of minimum variance. Moreover, an application of 
energy statistics to clustering in Euclidean spaces was 
considered \citep{Kgroups}.  
We refer to \citep{Szkely2013} for an overview.

\citet{Lyons} generalized
distance covariance from Euclidean 
to arbitrary metric spaces of strong negative type. Furthermore, 
the missing link between energy distance based tests and kernel 
based tests has 
been recently resolved by \citet{Sejdinovic2013}, where a unifying framework
establishing an equivalence between generalized energy distances to maximum
mean discrepancies (MMD), which are distances between embeddings of 
distributions in reproducing kernel Hilbert spaces (RKHS), was established. 
This equivalence immediately relates energy statistics to
kernel methods often used in machine learning, and form the basis 
of our approach in this paper.

@@

To account for nonlinearities, kernel methods were introduced 
\citep{Smola,Girolami}. A mercer kernel \citep{Mercer} is used to implicitly
map data points to a RKHS, then clustering can be performed in the associated
Hilbert space by exploiting its inner product. However, 
the kernel choice remains 
the biggest challenge since there is no principled theory to construct a kernel
for a given dataset, and usually kernels introduce hyperparameters that 
need to be carefully chosen.
The well-known kernel $k$-means optimization problem is nothing but $k$-means 
in the feature space \citep{Girolami}. Furthermore, kernel $k$-means algorithm
\citep{Dhillon2,Dhillon} is still based on Loyd's heuristic 
of grouping points that are closer to a cluster center, now
in the feature space. 
We refer the reader to \citet{Filippone} for a survey of clustering
methods.

Although clustering from energy statistics in Euclidean spaces was considered
in \citet{Kgroups}, the precise optimization problem behind this approach
remains elusive, as well as the connection with kernel methods.
The main theoretical contribution of this paper is to fill this gap.
Since the statistical potential energy is minimum when
distributions are equal, the principle behind clustering is to maximize 
the statistical energy,  enforcing probability distributions associated to 
each cluster to be different from one another. We provide a precise 
mathematical formulation to this statement, leading to a quadratically 
constrained quadratic program (QCQP) in the associated RKHS. Our results
immediately establish the connection between energy statistics based
clustering and kernel methods. Moreover, we show that that this 
QCQP is equivalent to kernel $k$-means optimization problem.

Our main 
algorithmic contribution is to use Hartigan's method \citep{Hartigan} to 
find local solutions of the above mentioned QCQP, which is NP-hard in general.
Hartigan's method was also used by \citet{Kgroups}, but without any
connection to kernels. More importantly, the 
advantages of Hartigan's
over Lloyd's method was already demonstrated 
in some simple settings by
\citet{Telgarsky,Slonin}, but apparently this method did not receive 
the deserved attention by the statistics and machine learning communities. 
To the 
best of our knowledge, Hartigan's method was not previously 
employed together with kernel methods. 
Here we provide a fully kernel based Hartigan's algorithm for clustering,
where the kernel is fixed by energy statistics. 
We make clear the advantages of this proposal versus 
Lloyd's method, which kernel $k$-means is based upon and will also be used 
to solve our QCQP. We show that both algorithms  have the same
time complexity, however, Hartigan's method in kernel spaces is superior. 
Furthermore, in the examples considered in this paper, it 
also provides equal or better performance than spectral clustering,
which in fact solves a relaxed version of our QCQP.
Our numerical results provide compelling evidence that 
Hartigan's method applied to energy statistics based clustering
is more accurate and robust than 
kernel $k$-means. Moreover, we illustrate the
flexibility of energy clustering,  
showing that it is able to perform accurately on data coming from 
very different distributions, contrary to $k$-means and GMM, for instance.
More specifically, the proposed method performs 
closely to $k$-means and GMM on normally distributed data, however,
it is significantly better on other settings. 
Its superiority in high dimensions is striking, being more accurate 
than $k$-means and GMM even on Gaussian settings.
We illustrate the applicability of the proposed method on a real dataset
obtained by neuroscientists experts, which describes 
shapes of neural synapses.

@@

In this section we introduce the main concepts from (generalized) energy
statistics \citep{Szkely2013,Lyons} and its relation to 
RKHS \citep{Sejdinovic2013}.

Consider random vectors $X_i$ ($i=1,\dotsc,n$) living in an arbitrary space
$\mathcal{X}$ of \emph{negative type}, which means that $\mathcal{X}$
is endowed with a \emph{semimetric} 
$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$ satisfying
$\sum_{i,j=1}^n c_i c_j \rho(X_i, X_j) \le 0$,
where $c_i \in \mathbb{R}$ and
$\sum_{i=1}^n c_i = 0$. 
Let $X,X' \stackrel{iid}{\sim} P$ and 
$Y,Y' \stackrel{iid}{\sim} Q$, where $P$ and $Q$ are cumulative
distribution functions with finite first moments, and 
$X,X',Y,Y' \in \mathcal{X}$. 
The \emph{generalized energy distance} between $P$ and $Q$ is
given by 
\begin{equation}
\label{eq:energy3}
\Energy(P, Q) \equiv 2 \E \rho(X,Y) - \E \rho(X, X') - \E \rho(Y,Y').
\end{equation} 
This quantity is nonnegative, $\Energy(P,Q) \ge 0$, and 
$\Energy^{1/2}$ is
a metric on the space of distributions. Energy distance provides
a characterization of equality of distributions.
For instance, the \emph{standard
energy distance} 
in Euclidean spaces 
introduced by \citet{Szkely2013} 
uses
the semimetric
\begin{equation}
\label{eq:rho_standard}
\rho_\alpha(X,Y) = \| X - Y\|^\alpha
\end{equation} 
where $0< \alpha \le 2$ and $\| \cdot \|$ is the
Euclidean norm in $\mathcal{X}=\mathbb{R}^D$. In this case, $\Energy(P,Q)$
is rotationally invariant. For $0<\alpha<2$ we have
$\Energy(P,Q) = 0$ if and only if $P=Q$. However, for $\alpha=2$
we get $\Energy(P,Q) = 2\| \E(X) - \E(Y) \|^2$, thus
$\Energy(P,Q)=0$ does 
not imply equality of distributions but only equality
of the means.

Consider a sample $\mathbb{X} = \{ x_1,\dotsc, x_n \}$ 
from $k$ unknown distributions $\{ P_j \}_{j=1}^k$,
where $x_i \in \mathcal{X}$.
Let $\mathbb{X} = \bigcup_{j=1}^k \C_j$ be a disjoint
partition, i.e. $\C_i \cap \C_j = \emptyset$.
Each expectation in the generalized energy distance
can be empirically estimated with the aid of the
function
\begin{equation}
\label{eq:g_def}
g (\C_i, \C_j) \equiv 
\dfrac{1}{n_i n_j}
\sum_{x \in \C_i} 
\sum_{y \in \C_j} \rho(x, y) ,
\end{equation}
where $n_i = |\C_i|$ denotes the number of points in partition
$\C_i$. 
Define the \emph{within energy dispersion} as
\begin{equation}
\label{eq:within}
W \equiv
\sum_{j=1}^{k} \dfrac{n_j}{2} g(\C_j, \C_j),
\end{equation}
and the \emph{between-sample energy statistic} as
\begin{equation}
\label{eq:between}
S \equiv \hspace{-1em}
\sum_{1 \le  i < j \le k } \hspace{-.5em} \dfrac{n_i n_{j}}{2 n} \left[
2 g(\C_i, \C_j) - 
g(\C_i, \C_i) - 
g(\C_j, \C_j)
\right],
\end{equation}
where $n = \sum_{j=1}^k n_j$.
A given point $x_i$ belongs to partition $\C_j$
if and only if $x_i \sim P_j$. 
The quantity $S$ defined above is
a test statistic for equality of distributions
\citep{Szkely2013}.
When the sample size is large enough, $n\to \infty$,
under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$ we have that
$S\to 0$, 
and under
the alternative hypothesis $H_1: P_\ell \ne P_j$ for at least two $\ell\ne j$, 
we have that $S \to \infty$.

Let $\HH_\kk$ be a Hilbert space of real-valued functions
over $\mathcal{X}$ with an associated kernel
$\kk : \mathcal{X} \times \mathcal{X} \to 
\mathbb{R}$, which is a symmetric and positive definite function.
%, i.e. 
%$\kk(x_i,x_j) = \kk(x_j,x_i)$ and 
%$\sum_{i,j=1}^n c_i c_j \kk(x_i, x_j) \ge 0$, or equivalently,
%if $G$ is the Gram matrix with entries $G_{ij} = \kk(x_i,x_j)$
%then $G = G^\top$ and $v^\top G v \ge 0$ for any $v \in \mathbb{R}^n$.
For every $x \in \mathcal{X}$ there exists 
$h_x \equiv \kk(\cdot,x) \in \Hk$ such that  $\langle h_x, f \rangle = f(x)$
for any function $f \in \Hk$. Thus, 
$\langle h_x, h_y \rangle = \kk(x,y)$.
Conversely, 
for every symmetric
positive definite function $\kk: \mathcal{X}\times \mathcal{X} \to
\mathbb{R}$ there is a Hilbert space $\Hk$ with reproducing
kernel $\kk$, with a 
\emph{feature map} $\varphi: x \mapsto \kkk_x \in \Hk$ such
that $\langle \varphi(x), \varphi(y) \rangle = \kk(x, y)$
\citep{Aronszajn}.

Define the embedding of a probability measure
$P \mapsto \kkk_P \in \Hk$ through
$\kkk_P \equiv \int \kk( \, \cdot \,, x)  d P(x)$. 
The distance between two probability measures, 
called maximum mean discrepancy (MMD), is thus given by
\begin{equation}
\label{eq:mmd}
\gamma_\kk(P,Q) \equiv \| \kkk_P - \kkk_Q \|_{\Hk},
\end{equation}
which can also be written as \citep{Gretton2012}
\begin{equation}\label{eq:mmd2}
\gamma_\kk^2(P,Q) = \E \kk(X,X') + \E \kk(Y,Y') - 2 \E \kk(X, Y)
\end{equation}
where $X,X' \stackrel{iid}{\sim} P$ and $Y,Y'\stackrel{iid}{\sim} Q$.
The equality between \eqref{eq:mmd} and \eqref{eq:mmd2}
gives $\langle \kkk_P, \kkk_Q \rangle_{\Hk} = \E \, \kk(X, Y)$.
%thus in practice we can estimate the inner product between  
%embedded distributions by averaging the kernel function over sampled data.

The following important result of 
\citet{Berg1984}
shows that semimetrics of negative
type and symmetric positive definite kernels are closely related. 
Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$,
and $x_0 \in \mathcal{X}$ be an arbitrary but fixed point.
Define
\begin{equation}
\label{eq:kernel_semimetric}
\kk(x,y) \equiv 
\tfrac{1}{2} \left[  \rho(x,x_0) + \rho(y,x_0) - \rho(x,y)\right].
\end{equation}
Then, it can be shown that 
$\kk$ is positive definite if and only if $\rho$ is a semimetric
of negative type.
We thus have a family of kernels, one for each choice of $x_0$. Conversely,
as shown by \citet{Sejdinovic2013},
if $\rho$ is a semimetric of negative type and $\kk$ is a kernel in this
family, then 
\begin{equation}
\label{eq:gen_kernel}
\begin{split}
\rho(x,y) &= \kk(x,x) + \kk(y,y) -2\kk(x,y) \\
&=  \| \kkk_x - \kkk_y \|^2_{\Hk}
\end{split}
\end{equation}
and the canonical feature map 
$\varphi: x \mapsto \kkk_x$ is injective. 
When the above conditions are met we say that the kernel $\kk$ 
generates the semimetric $\rho$. 
If two different kernels generate the same $\rho$ they are
said to be equivalent kernels.
Now we can state the equivalence between energy distance and
inner products on RKHS, which is one of the main results of
\citet{Sejdinovic2013}. If $\rho$ is a semimetric
of negative type and $\kk$ a kernel that generates $\rho$, then
replacing \eqref{eq:gen_kernel} into
\eqref{eq:energy3}, and using \eqref{eq:mmd2}, yields
\begin{align}
\tfrac{1}{2}\Energy(P, Q) &= 
\E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y) \nonumber \\ 
&= \gamma_\kk^2(P,Q) .
\label{eq:Erho}
\end{align}
Therefore, we can compute the energy distance using the inner product
of the associated Hilbert space $\Hk$.

@@

This section contains our main theoretical contributions, where 
we formulate an optimization problem for clustering 
based on energy statistics in the associated RKHS.
The proofs are contained in supplementary material.

Due to the energy test statistic for equality of distributions previously
discussed, the obvious
criterion for clustering data is to 
maximize $S$, defined in \eqref{eq:between}, which makes 
each cluster as different
as possible from the other ones.
In other words, given a set of points coming from different probability
distributions, the test statistic $S$ should attain a maximum when 
each point is correctly
classified as belonging to the cluster associated to its probability
distribution.
The following 
straightforward result
shows that maximizing $S$ is, however, equivalent to minimizing
$W$, which has a more convenient form.

\begin{lemma}
\label{th:minimize}
Let $\mathbb{X} = \{x_1,\dotsc,x_n\}$, where each data point
$x_i$ lives in a space $\mathcal{X}$ of negative type. 
For a fixed integer $k$,
the partition
$\mathbb{X} = \bigcup_{j=1}^k \C^\star_j$, where 
$\C^\star_i \cap C^\star_j = \emptyset$ for
all $i\ne j$, maximizes the between-sample statistic $S$, defined
in \eqref{eq:between}, if and only if
\begin{equation}
\label{eq:minimize}
\{C_1^\star, \dotsc,\C_k^\star\} = \argmin_{\{\C_1,\dotsc,C_k\}  } W(
\C_1, \dotsc, \C_k),
\end{equation}
where the within energy dispersion $W$ is defined by \eqref{eq:within}.
\end{lemma}

In the particular Euclidean case, 
the optimization problem \eqref{eq:minimize} based on
energy statistics was already proposed in \citet{Kgroups}. However, it is
important to note that this is equivalent to maximizing $S$,
which is the test statistic for equality of distributions.

The clustering
problem as stated in the
form \eqref{eq:minimize} makes
the relation with kernels and other clustering methods obscure.
In the following, we demonstrate what is the explicit optimization 
problem behind 
\eqref{eq:minimize} in the associated RKHS, directly 
establishing the connection with kernel methods commonly used in machine
learning.
